---
title: "Contributing to palaeoverse: structure and standards"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contributing to palaeoverse: structure and standards}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE}
# jarl-ignore-file unreachable_code: https://github.com/etiennebacher/jarl/issues/716
```

# Introduction

The `palaeoverse` R package is a community-driven software library providing generic tools for palaeobiological analysis. The core principles of palaeoverse are to: (1) streamline analyses, (2) enhance code readability, and (3) improve reproducibility of results.

This document describes the essential structure and conventions of the `palaeoverse` R package. Naturally, there are always disagreements regarding best practices and conventions, and `palaeoverse` is no exception. Despite this, all the essentials in `palaeoverse` are encouraged to make the lives of both the developer, and the user, easier. It is worth noting that the core structure and conventions adopted in `palaeoverse` are heavily influenced by Hadley Wickham and Jenny Bryan's [R Packages](https://r-pkgs.org/index.html) and the [tidyverse style guide](https://style.tidyverse.org/index.html), which is currently also Google's guide.  


# Files

File names should always be concise, meaningful and end in `'.R'`. Avoid using special characters whenever possible (i.e. stick to letters and numbers), and use `'_'` or `'-'` instead of spaces in file names. The use of lowercase is also strongly encouraged, and **never** have file names that only differ in capitalization. **Note:** the preferred separator is `'_'` to consist with the 'lowercase snake_case' convention for developing functions (more on that later!).  
<br>  

```{r names, include=TRUE, eval = FALSE}
# Bad
Sup3rAw3sumFuncti0n.r

# Better
super_awesome_function.R

# Best
time_bins.R
```

# R style guide

> "Good coding style is like correct punctuation: you can manage without it, butitsuremakesthingseasiertoread." -- tidyverse style guide

Below is a list of conventions for code, comments, and other forms of text in the package.

## Code formatting

Code formatting is automatically enforced by [Air](https://posit-dev.github.io/air/) and checked in every pull request. We recommend installing it and [setting it up](https://posit-dev.github.io/air/editors.html) in your preferred editor to enable features such as format-on-save.

This code formatter ensures that the following rules are respected:

* assignments use arrows (`<-`) rather than equals sign (`=`)

* indentation and spacing is consistent and favours readability throughout the code 

* in rare cases, you may want to [disable formatting](https://posit-dev.github.io/air/formatter.html#disabling-formatting) on specific pieces of code, but this should be the exception and not the rule. 

## Comments

Comments should be used for two purposes only:

1. explain tricky code
1. explain why the code is written as it is

In particular, except in complicated cases (cf point 1), comments should explain the "why" and not the "what" or "how".
For example, none of the following comments are useful because they just duplicate the information about what the code does:

```{r eval = FALSE}
f <- function(x) {
  # Only keep observations where y != 1
  x <- subset(x, y != 1)

  # Throw an error if there are missing values
  if (anyNA(x)) {
    stop("x must not contain missing values")
  }

  # Return x
  return(x)
}
```

Better comments could be for example:

```{r eval = FALSE}
f <- function(x) {
  # Observations where y != 1 can cause a crash later on.
  # https://github.com/palaeoverse/palaeoverse/issues/xxx
  x <- subset(x, y != 1)

  if (anyNA(x)) {
    stop("x must not contain missing values")
  }

  return(x)
}
```

Here, the first comment explains why we subset the data and links to an existing issue so that future readers of this code know exactly where to get more details. 
The other two comments can be removed since the code is self-explanatory.

## Object names

Use object names that are short but relevant to their contents, ideally in 'lowercase snake_case':

```{r eval = FALSE}
# Good 
lat_range

# Bad
latRange
latrange
lat.range
```

## Other text

Use plain English (British spelling preferred, e.g. `behaviour` instead of `behavior`) for your code and documentation.

# Data

Often we might want to include data into palaeoverse. This might come in the form of example datasets for testing functions (e.g. fossil occurrences), reference datasets such as the Geological Timescale 2020, or even data that is fundamental for a function to run. In palaeoverse's structure, we currently recognize four main ways of including data in the package depending on its usage.

- Raw data
- Internal data
- Exported data
- External data (preferred option)

However, for the sake of file size efficiency, please only include data and variables in your files which will be used by your function. **Note:** in order to be released on CRAN, R packages must be less than 5 MB in size. Please only include data that is absolutely vital for your function. The preferred solution to including data into palaeoverse functions is calling the data via an API or download URL (i.e. external data).

## Raw data

Raw data should always be included in `inst/extdata`. If you want to include cleaned/processed data in `data/`, it is generally a good idea to include the code used to process the raw data. If you ever need to reproduce or update your cleaned data, this will save you precious time. The code for processing your data should be included in `data-raw/`. Strictly speaking, raw data does not need to be documented. However, it is a good idea to include the original source and version (including a download date) in your code.

## Internal data

Data you do not wish to directly make available to the user should be saved as `R/sysdata.rda`. This is the best option for pre-computed data tables that are needed for a function to run. Strictly speaking internal data does not need to be documented. However, it is a good idea to document the internal data in the function documentation. 

## Exported data

Package data you wish to make available to the user should be stored in `data/`. Each file in this directory should be either a `'.rda'` or `'.RData'`. This file type is fast, small and explicit. The most appropriate way to include exported data is to use `usethis::use_data()`.

When using large datasets, we want to ensure that our files are not bloated and taking up too much space on our users' machines. As such, you may want to experiment with the compression settings in `usethis::use_data()`. Generally, `xz` and `gzip` can create smaller files than the default `bzip2`. You can also implement several 'hacks' to generate smaller files which you may want to consider for large datasets (depending on whether your data is sensitive to such changes). Data with many decimal places consume a lot of memory, consider how many significant figures are relevant for your data, and `round()` accordingly. You can also experiment with your file size by multiplying your data by X (e.g. 1,000) to remove decimal places altogether. **Note:** Remember to undo any transformations when calling or working with the data.

## External data

External data is the preferred approach for contributors to include data into `palaeoverse` and will be required in almost all cases. This is to ensure that the package does not become unnecessary bloated to all users when the data can just be called by a download link. Currently, external data are stored on a GitHub repo (reconstruction files for `palaeorotate`). It is likely that in the future this will be migrated elsewhere as GitHub is not a data repository (hopefully a dedicated server when funding allows!). If you wish to include external data, please get in touch with one of the palaeoverse developers, and we can discuss what might be the best solution.

```{r download, include = TRUE, eval = FALSE}
#generate temp directory
files <- tempdir()
#download files
download.file(
  url = "www.goo.com",
  destfile = paste0(files, "/mgoo.csv"),
  mode = "wb"
)
#run some kind function using the download
#REMEMBER: remove downloaded files
unlink(x = paste0(files, "/", list.files(files)))
```

## Data documentation

Objects in `data/` are always exported by default, and should be documented accordingly. In order to properly document data, you must document the name of the dataset and save it in `R/data`. For example, the documentation block used to document GTS2020 is saved as `R/data.R` and is similar to the following (simplified here): 

```{r document, include = TRUE, eval = FALSE}
#' Geological Time Scale 2020
#'
#' A dataset of the Geological Time Scale 2020. Age data from:
#'  \url{https://stratigraphy.org/chart/}.
#' Supplementary information is also included in the dataset for
#' plotting functionality (e.g. GTS2020 colour scheme).
#'
#' @format A data frame with 189 rows and 20 variables:
#' \describe{
#'   \item{index}{Index number for the order of all intervals in the dataset}
#'   \item{stage_number}{Index number for stages}
#'   \item{series_number}{Index number for series}
#'   \item{system_number}{Index number for system}
#'   \item{interval_name}{Names of intervals in the dataset}
#'   ...
#' }
#' @section References:
#' Gradstein, F.M., Ogg, J.G., Schmitz, M.D. and Ogg, G.M. eds. (2020).
#' Geologic time scale 2020. Elsevier.
#' @source Compiled by Lewis A. Jones. See item descriptions for details.
"GTS2020"
```

## What does it mean to document your data?

Documenting your data is to provide a thorough description of the data and any information relevant to understanding it. Good documentation is key to reproducible science, and will also help us to ensure that we acknowledge all data collators who have provided data for the `palaeoverse` package. When providing us with datasets, please give the following information:

**Author(s):** Who collected the data, and prepared its current format? Please provide citations if relevant. Have the authors given permission for the dataset to be included in the `palaeoverse` package? 

**Description:** Brief description of the dataset.

**Provenance:** When and how was the data collected? When was the dataset finalised in its current form?

**Size:** Please state the full, uncompressed size of the file.

**Variables:** Describe each of the columns in your dataset, providing as much information as you can on the full name of the variable, data type (e.g. continuous, discrete, categoric, etc.), units, and how it was collected.

# Functions

## Conventions

Functions are saved as `'.R'` files in the `'R'` folder. The name of the file needs to correspond to the function, e.g. the file `time_bins.R` contains the function `time_bins()`. Function names and arguments should be informative and aim to keep in line with available functions in `palaeoverse`. For example, all functions that bin data, whether it be by time or space, start with `bin_`, whereas taxonomic-related functions start with `tax_` (e.g. `tax_unique`, `tax_check`). Where possible, argument names should also consist with functions in the package. For example, if a function requires an occurrence dataframe as input, the argument name should be `occdf`. It is difficult to give a complete static summary of all the conventions as `palaeoverse` will undoubtedly evolve over time. We also wish to remain flexible for contributors, provided that it does not compromise the user-friendliness of the package. We welcome contributors to check through the source code of `palaeoverse` for function examples.

## Documentation
For documentation, we use [roxygen2](https://roxygen2.r-lib.org). The title, a brief description, and every argument (including the input class and default input) and the output of the function need to be documented in `roxygen2` style, for example:
```{r function, include = TRUE, eval = FALSE}
#' An exemplary function
#' This function is used to demonstrate the documentation of functions.
#' @param example \code{character}. Arguments are the function inputs.
#' @param another_example \code{logical}. All arguments need to be documented.
#' @return A \code{list} is returned as output in this example function.
#' @details Describe more details if necessary, and list sources if applicable.
#' @section Developer:
#' Your name
#' @section Reviewer:
#' Name(s)
#' @examples
#' #Show off the example function
#' example_function(example = “documentation”, another_example = TRUE)
#' @export
```
Add the `'@export'` namespace tag to make the function available.

To get started with `roxygen2`, set your working directory to the package directory, or to the directory where you store your function as a `'.R'` file. The R command `devtools::document()` creates a 'man' folder in the directory, which contains a `'.Rd'` file corresponding to your documented function. Opening that file in RStudio, you can create a preview to see what the documentation looks like. After you have implemented changes, rebuild your documentation file with Ctrl+Shift+B or `devtools::document()`.

## Input checking

All functions should start with at least one input check.

`palaeoverse` uses dedicated functions from the `rlang` package, e.g. `rlang::check_bool()` to ensure that a parameter is `TRUE` or `FALSE` and generate a good error message otherwise. It also contains custom functions to check inputs, e.g. `check_class()` to ensure that the column of a dataset has the right class. You can see the definition of these functions in `R/utils-rlang.R`, and you can explore how they are used in other `palaeoverse` functions.

## Errors and warnings

Many error messages are handled by the check functions mentioned in the previous section, but you may also want to throw an error or a warning with a custom message. `palaeoverse` uses the `cli` package to generate these messages: use `cli::cli_abort()` for an error (instead of `stop()`) and `cli::cli_warn()` for a warning (instead of `warning()`).

These messages must follow the rules detailed at the top of `R/utils-rlang.R`, e.g. column names must be wrapped in double quotation marks and argument names must be wrapped in backticks. Here's an example of how to do that with `cli`:

```{r eval = TRUE, error = TRUE}
# Suppose that we are in a function and that the column "foo" from the input data
# "dat" has duplicated values while it shouldn't.
cli::cli_abort("Column {.val foo} in {.arg dat} must not have duplicated values.")
```

## Tests
Testing is perhaps one of the most important parts of developing a function. If you have not already, we recommend reading through Chapter 12–Testing of ['R packages'](https://r-pkgs.org/testing-basics.html#test-organisation) before continuing here as `palaeoverse` follows this guidance.

Before a function can be added to `palaeoverse`, it needs to go through formal testing. This is required as hopefully your function will be very popular, and we need to ensure that it behaves as expected to avoid any issues. To do so, we make use of the R package `testthat`.

The initial setup for function testing with `testthat` is already established in palaeoverse, under the directory `'tests/testthat/'`. The organisation of test files must match that of `'R/'` files in `palaeoverse`. For example, the function `time_bins()` is saved as `time_bins.R` in the `R/` directory, and has an associated test file of `'test-time_bins.R'` in the `'tests/testthat'` directory. This ensures that associated function testing is clear.

**Tip:** the `usethis` package provides a helpful pair of functions for creating/alternating between these files:

- `usethis::use_r()`
- `usethis::use_test()`

Make sure to create enough tests within your `.R` file to cover all of the possible variants of a function. This includes creating tests that cover most or all optional arguments and the majority of options for those arguments (and the required arguments). Remember, even if you personally would not use a function for a particular reason, you must attempt to cover the majority of edge cases that may arise that are allowed by the function.

Related tests should be bundled within `test_that()` calls combined with strings of text to identify the broad reason for each bundle of tests (e.g., testing a function works with a particular type of data). Finally, if tests rely on data or packages outside of the `palaeoverse` community must have, they should be skipped if those data or packages are not available.

Ultimately, we aim to have >90% code coverage (95% preferred), which means 90% of the lines of code in our codebase should be tested by at least one test. Pushing code to GitHub will trigger a code coverage check, which will alert you as to whether you need to write more tests.

```{r test, include = TRUE, eval = FALSE}
# Test expected outcome is equal to actual outcome
test_that("time_bins() works", {
  expect_equal(nrow(time_bins(interval = c("Maastrichtian"))), 1)
})
```

# Contributing to palaeoverse

At `palaeoverse` we have adopted a set of [structures and standards](https://palaeoverse.palaeoverse.org/articles/structure-and-standards.html) to follow for contributing to the development of `palaeoverse`. If you would like to contribute to the `palaeoverse` toolkit, we strongly advise reading this document first. If you plan to contribute a function to `palaeoverse`, you should first raise an [issue](https://github.com/palaeoverse/palaeoverse/issues) via the GitHub repository (see below). This way the development team can assess whether the function is suitable or needed in the `palaeoverse` toolkit prior to submission.

## Git and GitHub
We use git via Github (under the [palaeoverse GitHub umbrella](https://github.com/palaeoverse)) to manage our R code and data. If you are not familiar with these tools, there are some excellent free resources available online:

- [Setting up git within RStudio (short)](https://github.com/llendway/github_for_collaboration/blob/master/github_for_collaboration.md)
- [Collaborating in R with git and GitHub (50 min. youtube video)](https://www.youtube.com/watch?v=megZYkCLMA4)
- [Git and GitHub and R packages (medium length)](https://rfortherestofus.com/2021/02/how-to-use-git-github-with-r/)
- [A comprehensive guide to git and R (very long)](https://happygitwithr.com/)

## How to contribute

### Minor changes

You can fix typos, spelling mistakes, or grammatical errors in the documentation directly on GitHub, provided it is done so in the source file. This means you’ll need to edit roxygen2 comments in the `.R` file, not the `.Rd` file.

### Substantial changes

If you would like to make a substantial change, you should first file an issue and make sure someone from the development team agrees that it’s needed. If you’ve found a bug, please file an issue that illustrates the bug with a reproducible example.

### Pull request process

You (the contributor) should clone the desired repository (i.e. [the palaeoverse R package](https://github.com/palaeoverse/palaeoverse)) to your personal computer. Before changes are made, you should switch to a new git branch (i.e., not the main branch). When your changes are complete, you can submit your changes for merging via a [pull request](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/proposing-changes-to-your-work-with-pull-requests/about-pull-requests) (“PR”) on GitHub. Note that a complete pull request should include a succinct description ([see function template](https://github.com/palaeoverse/palaeoverse/blob/main/pull_request_template.md)) of what the code changes do, proper documentation (via [roxygen2](https://roxygen2.r-lib.org)), and unit tests (via `testthat`). Only the description is required for the initial pull request and code review (see below), but pull requests will not be merged until they contain complete documentation and tests.

If you are not comfortable with git/GitHub, you can reach out to one of the core developers ([see collaborators](https://palaeoverse.palaeoverse.org/authors.html)) via email and they can make a pull request on your behalf. However, you will be expected to respond to any reviewer comments on GitHub (see below).

If you don't feel comfortable implementing changes yourself, you can submit a bug report or feature request as a GitHub issue in the proper repository (e.g. [palaeoverse issues](https://github.com/palaeoverse/palaeoverse/issues)).

## Code review
All pull requests must be reviewed by two core developers of `palaeoverse` ([see collaborators](https://palaeoverse.palaeoverse.org/authors.html)) before merging. The review process will ensure that contributions 1) meet the standards and expectations as described above, 2) successfully perform the functions that they claim to perform, and 3) don't break any other parts of the codebase.

Submitting a pull request to the `palaeoverse` package will automatically initiate an [R CMD check](https://r-pkgs.org/check.html), [lintr check](https://lintr.r-lib.org/index.html), and [test coverage check](https://covr.r-lib.org/) via GitHub Actions. While these checks will conduct some automatic review to ensure the package has not been broken by the new code and that the code matches the style guide (see above), a manual review is still required before the pull request can be merged.

Reviewers may have questions while reviewing your pull request. You are expected to respond to any of these questions via GitHub. If fixes and/or changes are required, you are expected to make these changes. If the required changes are minor enough, reviewers may make them for you, but this should not be expected. If you have any questions or lack the background to make the required changes, you should work with the reviewer to determine a plan of attack.

</div>
